Icicle: a single-pass query language

### 2015-12-05 2016icicle/section/02-Source.tex
related work: lustre
ok
changes

The first example defines @sum@ as a fold over the input data.
The initial value for the fold is $0$, followed by the previous value @s@ added to the current value @v@.
\begin{code}
sum v
 =  fold s = 0; s+v
 ~> s
\end{code}

Next, we can define @count@ and mean in terms of this sum.
\begin{code}
count
 =  sum 1

mean v
 =  sum v / count
\end{code}

We can use @filter@ to count only the days where the close price is higher than the open.
\begin{code}
gap
 =  filter close > open
 ~> count

proportion
 =  gap / count
\end{code}

Finally, we can use @group@ to put into buckets based on the percentage of growth.
Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
\begin{code}
growthPercent
 =  int ((close - open) / open * 100)

growthBy
 =  let total = count
 ~> group growthPercent
 ~> count / total
\end{code}

@@

@@

>>>>

The first example defines $\mi{sum}$ as a fold over the input data.
The initial value for the fold is $0$, followed by the previous value $s$ added to the current value $v$.

\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{sum}~v$    \\
\> $=$  \> $@fold@~s~=~0~@then@~s~+~v$ \\
\> @in@ \> $s$ \\
\end{tabbing}

Next, we can define $\mi{count}$ and $\mi{mean}$ in terms of this sum.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{count}$                                        \\
 \> $=$  \> $\mi{sum}~1$                            \\
                                                    \\
$\mi{mean}~v$                                       \\
 \> $=$  \> $\mi{sum}~v~/~\mi{count}$               \\
\end{tabbing}

We can use @filter@ to count only the days where the close price is higher than the open.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{gap}$                                          \\
 \> $=$  \> $@filter@~\mi{close}~>~\mi{open}$       \\
 \> @in@ \> $\mi{count}$                            \\
                                                    \\
$\mi{proportion}$                                   \\
 \> $=$  \> $\mi{gap}~/~\mi{count}$
\end{tabbing}

Finally, we can use @group@ to put into buckets based on the percentage of growth.
Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{growthPercent}$                                        \\
 \> $=$  \> $((\mi{close}-\mi{open}) \times 100 ) / \mi{open}$  \\
                                                            \\
$\mi{growthBy}$                                             \\
 \> $=$  \> $@let@~\mi{total} = \mi{count}$                 \\
 \> @in@ \> $@group@~\mi{growthPercent}$                    \\
 \> @in@ \> $\mi{count} / \mi{total}$                    \\
\end{tabbing}

@@

\TODO{Talk about scan}

We formally define the grammar of the language in figure~\ref{fig:source:grammar}.
Note that general application is not allowed: arguments can only be applied to primitives and named functions.
Similarly, functions and primitives must be fully applied.
This, combined with the lack of lambda construction, means that higher order functions are outlawed.
Outlawing higher order functions is somewhat unsatisfying, but simplifies the typing rules and makes it easier to support our performance guarantees.

While the examples shown allow user defined functions, there is no recursion allowed.
This means that all function definitions can be inlined into the user query with guaranteed termination.
When converting to the intermediate language, we assume that this inlining has occurred.

\TODO{Should we go through the evaluation rules here, or after typechecking?}
If we can write the evaluation rules to be ``agnostic'' of the type, it would be ideal to put them here.

@@

\input{figures/Source-Eval.tex}

### 2015-12-05 2016icicle/section/03-Types.tex
ok
nothing much happening
changes

For (TcFold), the form of fold is $@fold@~v~@=@~x@;@~x'$ where $v$ is the name of the fold binding, $x$ is its initial value, and $x'$ computes the successor value, based on the previous value of $v$.
The initial expression $x$ must be @Pure@ of some type $\tau$, and the successor expression $x'$ is typechecked with the fold binding set to $@Element@~\tau$ in the environment.
The result of the successor expression must also have type $@Element@~\tau$.
This is because the current value of the fold is available to the successor expression, as well as the current elements, but the results of other folds are not available.
In the remaining context, the fold value is only available as an @Aggregate@.

For (TcLatestElement), if you have a $@latest@~n$ of @Element@s we fill a circular buffer with the last $n$ values.
Similarly, (TcLatestAggregate) uses a circular buffer, and then applies the aggregate to the last $n$ values.

@@

>>>>

For (TcFold), the form of fold is $@fold@~v~@=@~x~@then@~x'$ where $v$ is the name of the fold binding, $x$ is its initial value, and $x'$ computes the successor value, based on the previous value of $v$.
The initial expression $x$ must be @Pure@ of some type $\tau$, and the successor expression $x'$ is typechecked with the fold binding set to $@Element@~\tau$ in the environment.
The result of the successor expression must also have type $@Element@~\tau$.
This is because the current value of the fold is available to the successor expression, as well as the current elements, but the results of other folds are not available.
In the remaining context, the fold value is only available as an @Aggregate@.

@@

\input{figures/Source-Compile.tex}

### 2015-12-05 2016icicle/section/XX-Related-DataFlow.tex
changes

While this may be the desired behaviour in some cases, we believe this decision must be made explicit.
At the moment this must be hand-written as a fold, but we intend to add a new primitive called @running@ which peeks into a fold and uses its intermediate result.
\begin{code}
   filter value > running (mean value)
-> count
\end{code}

>>>>

While this may be the desired behaviour in some cases, we believe this decision must be made explicit.
At the moment this must be hand-written as a fold, but we intend to add a new primitive called @scan@ which peeks into a fold and uses its intermediate result.
\begin{code}
   filter value > scan (mean value)
-> count
\end{code}
